Equations & R Code

### 2017-05-09 manuscript/coexistence-stability_ms.tex
final edits for ELE

MacArthur (1955), Elton (1958), and even Darwin (Turnbull et al. 2013)
recognized the potential for compensatory dynamics among species to
stabilize ecosystem functioning in fluctuating environments. This idea
underlies the ``insurance hypothesis'' (Yachi \& Loreau 1999), which
\new{states that ecosystem variability, defined as the coefficient of variation of ecosystem biomass over time, should}
decrease with diversity because species respond dissimilarly to
environmental variation, broadening the range of conditions under which
the community maintains function (Loreau 2010). A variety of theoretical
models all predict a negative relationship between species richness and
ecosystem variability (Lehman \& Tilman 2000; Ives \& Hughes 2002;
Loreau \& de Mazancourt 2013), and experimental tests tend to support
such a prediction (Tilman et al. 2006; Hector et al. 2010).

However, the ability of biodiversity--ecosystem functioning (BEF)
experiments to accurately represent real-world dynamics is debated
(Eisenhauer et al. 2016; Wardle 2016). Much of the debate centers on the
fact that BEF experimental protocols
\new{do not allow species additions from the regional pool to offset species losses in local communities.}
Theoretical work on diversity--\new{ecosystem variability} relationships
typically suffers from the same limitation: it recognizes the role of
environmental variability in driving population fluctuations which
destabilize ecosystems, but ignores the potential for environmental
variability to promote species richness and thereby help stabilize
ecosystems (Loreau 2010, but see Chesson et al. 2001).

Fluctuating environmental conditions are an important ingredient for
stable species coexistence, both in theoretical models (Chesson 2000a;
Chesson et al. 2004) and in natural communities (Cáceres 1997;
Descamps-Julien \& Gonzalez 2005; Adler et al. 2006; Angert et al.
2009). Such ``fluctuation-dependent'' coexistence
\new{emerges most easily when} species have unique environmental
responses and environmental conditions vary so that each species
experiences both favorable and unfavorable conditions, preventing
competitive exclusion (Chesson 2000a).
\new{Chesson (2000) described the two fluctuation-dependent mechanisms: the storage effect and relative nonlinearity.
Both mechanisms operate when environmental variation favors different species at different times. 
Under the storage effect, this happens because species are competing for resources at different times (and escaping competition in unfavorable periods). 
Under relative nonlinearity, all species are competing for resources at the same time, but each species alters resource availability in a way that favors its competitors.
We describe these mechanisms in more detail below (see \textbf{Materials and Methods: Consumer-resource model}).}

@@

The opposing effects of environmental variability on ecosystem
variability might explain the mixed results from \new{observational}
studies on the diversity--\new{ecosystem variability} relationship.
Observational tests of the diversity--\new{ecosystem variability}
relationship, which require sampling across natural diversity gradients,
have yielded negative (Hautier et al. 2014), neutral (Valone \& Hoffman
2003; Cusson et al. 2015), and positive (Sasaki \& Lauenroth 2011)
relationships. In a meta-analysis of
diversity--\new{ecosystem variability} relationships, Jiang \& Pu (2009)
found no significant evidence for an effect of species richness on
ecosystem variability when restricting data to observational studies in
terrestrial ecosystems, perhaps because environmental variability varies
across natural diversity gradients, affecting both richness and
ecosystem variability. The idiosyncratic results of these observational
studies contrast with the consistent conclusions from
\new{experimental and theoretical work that ignore, or control, the}
feedbacks between variability and richness.

The gap between theoretical expectations and empirical results of
diversity--\new{ecosystem variability} relationships might reflect the
divergence of theory developed to explain species coexistence and theory
developed to explain diversity and ecosystem variability.
\new{In his thorough review of the topic, Loreau (2010) cautions that "one of the pieces of the stability jigsaw [puzzle] that is still missing here is the interconnection between community stability and the maintenance of species diversity due to temporal environmental variability."}
One reason these two disciplines have diverged is because they have
focused on different questions. Diversity--\new{ecosystem variability}
studies typically ask how ecosystem variability responds to different
levels of species richness at a given level of environmental variability
(reviewed in Kinzig et al. 2001; Loreau 2010), whereas coexistence
studies ask how species richness responds to different levels of
environmental variability (Chesson \& Warner 1981).

To reconcile these two perspectives, we extend theory on the
relationship between species richness and ecosystem variability to cases
in which species coexistence explicitly depends on environmental
fluctuations and species-specific responses to environmental conditions.
We focus on
\new{communities where coexistence is maintained by either the temporal storage effect or relative nonlinearity}
using a general consumer-resource model. We use the model to investigate
two questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Does the diversity--\new{ecosystem} variability relationship remain
  negative when species coexistence is maintained by the temporal
  storage effect or relative nonlinearity?
\item
  How does increasing environmental variability impact ecosystem
  variability when coexistence depends on the storage effect or relative
  nonlinearity?
\end{enumerate}

@@

We developed a semi-discrete consumer-resource model that allows
multiple species to coexist on one resource by either the storage effect
or relative nonlinearity. In our model, the consumer can be in one of
two-states: a dormant state \(D\) and a live state \(N\). The dormant
state could represent, for example, the seed bank of an annual plant
\new{or root biomass of a perennial plant}. Transitions between \(N\)
and \(D\) occur at discrete intervals between growing seasons, with
continuous-time consumer-resource dynamics between the discrete
transitions. Thus, our model is formulated as ``pulsed differential
equations'' (Pachepsky et al. 2008; Mailleret \& Lemesle 2009; Mordecai
et al. 2016).
\new{We refer to $\tau$ as growing seasons and each growing season is composed of $T$ daily time steps, indexed by $t$ ($t=1,2,3,\dots,T$).
For example, the notation $\tau(t)$ reads as: ``day $t$ within growing season $\tau$.''}

\new{At the beginning of growing season $\tau$ a season-specific fraction ($\gamma_{i,\tau}$) of dormant biomass is activated as living biomass such that}
\vspace{-3em}

@@

\noindent \new{where \emph{i} indexes each species and $\tau(0)$ denotes the beginning of growing season $\tau$.}
\new{Live biomass at the start of the growing season $\left(N_{i,\tau(0)}\right)$ then serves as the initial conditions for continuous-time}
consumer-resource dynamics that are modeled as two differential
equations: \vspace{-3em}

@@

\new{At the end of the growing season ($t=T$), a fraction ($\alpha_i$) of live biomass is stored as dormant biomass and a fraction of dormant biomass survives ($1-\eta_i$) to the next growing season, giving the following equation:}
\vspace{-3em}

@@

\noindent \new{where $\tau(T)$ denotes the end of growing season $\tau$.
We assume remaining live biomass $\left(N_{i,\tau(T)}(1-\alpha_i)\right)$ dies (i.e., this is not a closed system where all biomass must be in either $N$ or $D$ states).
We do not include extinction thresholds, or any other form of demographic stochasticity, under the assumption that we are working with abundant species with generous seed dispersal.}

We assume the resource pool is not replenished within a growing season.
Resource replenishment occurs between growing seasons, and the resource
pool (\emph{R}) at the start of the growing season is
\(R_{\tau(0)} = R^+\), where \(R^+\) is a random resource pulse drawn
from a lognormal distribution with mean \(\mu(R^+)\) and standard
deviation \(\sigma(R^+)\).
\new{Taken all together, we can combine equations 1 and 4 to define the discrete transitions between live and dormant biomass at the end of a growing season.
Thus, the initial conditions for each state ($D, N, R$) at the beginning of growing season $\tau+1$ are:}
\vspace{-3em}

@@

\noindent \new{where, as above, $\tau(T)$ denotes the end of growing season $\tau$ and $\tau(0) + 1$ denotes the beginning of growing season $\tau+1$.}
\new{The subscript (0) and superscript (200) indicates a lognormal distribution truncated at those values to avoid extreme resource pulses that cause computational problems.
We used the function \texttt{urlnorm} from the \texttt{Runuran} package}
(Leydold \& Hörmann 2015)
\new{to generate values from the truncated lognormal distribution.}
Model parameters and notation are described in Table 1.

\new{Our model does not include demographic stochasticity, which can lead to stochastic extinction for small populations as environmental  variability increases}
(Boyce 1992).
\new{Previous work has shown how demographic stochasticity and coexistence mechanisms can interact to create a weak "humped-shape" relationship between coexistence time and environmental variability}
(Adler \& Drake 2008),
\new{because environmental variability increases coexistence strength and the probability of stochastic extinction simultaneously.
We do not consider this potential effect here because our focus is on large populations that would most influence ecosystem functioning.}

\new{We limit our analysis to four-species communities because it is exceedingly difficult to get more than four species to coexist via relative nonlinearity without introducing another coexistence mechanism}
(Yuan \& Chesson 2015). For consistency, we also constrain our focus to
four species communities under the storage effect, but our conclusions
apply to more species-rich communities (see Supporting Information
section SI.2).\}

@@

For the storage effect to operate, we need species-specific responses to
environmental variability, density-dependent covariance between
environmental conditions and competition (\emph{EC} covariance), and
subadditive population growth (Chesson 1994, 2000b). If these conditions
are present, all species can increase when rare and coexistence is
stable. In the storage effect, rare species increase by escaping the
effects of \emph{EC} covariance. Common species will experience greater
than average competition (\emph{C}) in environment (\emph{E}) years that
are good for them because common species cannot avoid intraspecific
competition. However, a rare species can escape intraspecific
competition and has the potential to increase rapidly in a year when the
environment is good for them but bad for the common species. \emph{EC}
covariance emerges in our model because dormant-to-live transition rates
(\(\gamma\)) are species-specific and vary through time.
\new{In a high $\gamma$ year for a common species, resource uptake will be above average because combined population size will be above average.
In a year when $\gamma$ is high for rare species and low for common species, resource uptake will be below average because combined population size will be below average.}
Subadditive population growth buffers populations against large
population decreases in unfavorable years. It is included in our model
through a dormant stage with very low death rates, which limits large
population declines in bad \emph{E} years.

@@

To explore how fluctuation-dependent coexistence can affect the
diversity--ecosystem variability relationship, we simulated the model
with four species under two scenarios for each coexistence mechanism.
First, we allowed the variance of the environment to determine how many
species can coexist, akin to a community assembly experiment with a
species pool of four species.
\new{We simulated communities with all species initially present across a gradient of annual resource variability for relative nonlinearity (50 evenly-spaced values of $\sigma_R$ in the range [0, 1.2]) or environmental cue variability for the storage effect (100 evenly-spaced values of $\sigma_E^2$ in the range [0, 3]).}
Second, we chose parameter values that allowed coexistence of all four
species and then performed species removals
\new{at a single level of environmental variability}, akin to a
biodiversity--ecosystem function experiment. The two simulation
experiments correspond to (i) sampling ecosystem function across a
natural gradient of species richness and (ii) sampling ecosystem
function across diversity treatments within a site.
\new{We refer to the former as a "regional" relationship, and the latter as a "local" relationship.
But we do not attribute any particular area size to "region", it is simply any area over which a gradient of environmental variability exists.}

To understand how increasing environmental variability will impact
ecosystem variability when coexistence is fluctuation-dependent, we
simulated the model over a range of species pool sizes and environmental
cue or resource variability. For each size of species pool (1, 2, 3, or
4 species), we simulated the model at 15 evenly-spaced levels of
environmental cue (range = {[}0.1, 2{]}) for the storage effect and 25
evenly-spaced levels of resource variability (range = {[}0, 1.2{]}) for
relative nonlinearity. We also explored the influence of asymmetries in
species' competitive abilities and correlations in species'
environmental responses within the storage effect model.
\new{We created competitive hierarchies by making the live-to-dormant biomass fractions ($\alpha$s) unequal among species.
Small differences among values of $\alpha$ were needed to create competitive hierarchies because we chose a relatively constrained gradient of environmental cue variance.
Larger differences among values of $\alpha$ expand the region of coexistence farther along a gradient environmental cue variance.}

Under relative nonlinearity, species' resource response curves (Fig.
SI-5) reflect traits that determine the temporal variability of each
species' population growth. ``Stable'' species achieve maximum resource
uptake at low resource levels, but their maximum uptake rates are
modest. For these species, population responses to resource fluctuations
are buffered. ``Unstable'' species have very high maximum uptake rates,
which they only achieve when resource availability is high, leading to
large population fluctuations. The difference in the intrinsic stability
of these two kinds of species makes our simulations sensitive to initial
conditions. Therefore, we ran two sets of simulations for relative
nonlinearity: beginning with either stable or unstable species as a
reference point. For example, if species A is the most stable species
and species D is the least stable, we ran simulations where A then B
then C then D were \new{added to the initial pool of species}. We then
ran simulations with that order reversed.

All simulations were run for 5,000 growing seasons of 100 days each. We
averaged biomass over the growing season, and yearly values of
\new{live-state biomass} were used to calculate total community biomass
in each year. After discarding an initial 500 seasons to reduce
transient effects on our results, we calculated the coefficient of
variation (\emph{CV}) of summed species biomass through time, which
represents ecosystem variability, the inverse of ecosystem stability. We
calculated \new{realized} species richness as the number of species
whose average biomass was greater than 1 over the course of the
simulation.
\new{In some cases, realized species richness is less than number of species initialized for a simulation because of competitive exclusion.}

\new{For parameters that we did not vary, we chose values that would allow coexistence of all four species at some point along the environmental variability gradients we simulated.
Our focus is specifically on communities where fluctuation-dependent coexistence is operating, and making parameters increasingly asymmetric among species typically reduced coexistence strength or made coexistence impossible (Supporting Information section SI.3).
Changes in the absolute values of parameters also altered the strength of coexistence, but in no case did altering parameter values change our qualitative results and conclusions (Supporting Information section SI.3).}
Parameter values for specific results are given in figure captions.

Within-season dynamics were solved given initial conditions using the
package \texttt{deSolve} (Soetaert et al. 2010) in \texttt{R} (Team
2013). \texttt{R} code for our model function is in the Supporting
Information section SI.1. All model code has been deposited on Figshare
(\emph{link}) and is available on GitHub at
\url{http://github.com/atredennick/Coexistence-Stability}.

@@

When we allowed the variance of the environment to determine which of
four initial species coexisted, similar to a study across a natural
diversity gradient, we found a positive relationship between richness
and \new{ecosystem variability}, defined as the temporal \emph{CV} of
total community biomass (Fig. 2A,C). This was true for the storage
effect, where coexistence is maintained by fluctuating dormant-to-live
transition rates (\(\gamma\)), and for relative nonlinearity, where
coexistence is maintained by annual resource pulses. The relationship is
driven by the fact that increasing environmental variability increases
the strength of both coexistence mechanisms (Fig. SI-6). More variable
conditions promoted species richness, creating a positive relationship
between diversity and \new{ecosystem} variability.

When we performed species removals but held environmental variability at
a level that allows coexistence of all four species, similar to a
biodiversity--ecosystem functioning experiment, we found a negative
diversity--\new{ecosystem variability} relationship (Fig. 2B,D). Scatter
around the relationship was small for the storage effect because all
species have similar temporal variances. Regardless of species identity,
the presence of more species always stabilized ecosystem functioning
through portfolio effects. In contrast, scatter around the relationship
was larger for relative nonlinearity (Fig. 2D) because species with
different resource uptake curves had different population variances.
Depending on which species were present, two-species communities were
sometimes less variable than three-species communities.
\new{Furthermore, the slope of the relative nonlinearity diversity--ecosystem variability relationshp in Fig. 2D is sensitive to species' traits: the difference among species' resource uptake determines the spread of single-species communities along the y-axis.
This means that the relationship can become flat as species become more similar.}

\new{For the storage effect, total community \emph{CV} decreased with species richness at a given level of environmental variability because additional species reduced the temporal standard deviation due to portfolio effects (Fig. SI-7).
Mean biomass remained the same because all species had the same resource uptake functions, which was necessary to eliminate any potential effects of relative nonlinearity.
Portfolio effects under the storage effect remained strong in an eight-species community, where total community \emph{CV} saturated after addition of the fifth species (Fig. SI-1).
For relative nonlinearity, total community \emph{CV} decreased with species richness at a given level of environmental variability because additional species increased mean biomass (over-yielding) and, at higher richness (three to four species), reduced the temporal standard deviation (Fig. SI-7).
Mean biomass increased because some species had higher growth rates (Fig. SI-5), increasing total biomass.}

To understand how much species additions might stabilize ecosystem
functioning as environmental variability increases, we simulated our
model over a range of environmental variance and species pool sizes. For
both coexistence mechanisms, realized species richness increased with
environmental variability and, in some cases, increases in richness
completely offset the effect of moderate increases in environmental
variability on ecosystem variability (Fig. 3 and 4). More species rich
communities were less variable on average and, under the storage effect,
they increased in ecosystem \emph{CV} at a slower rate than communities
with fewer species (e.g., lower slopes in log-log space; Fig. SI-8).
\new{The buffering effect of species richness under the storage effect is also evident in Fig. 2A because the relationship between species richness and ecosystem \emph{CV} begins to saturate.
In fact, ecosystem \emph{CV} remains relatively constant past four species when species have independent responses to the environment ($\rho=0$; Fig. SI-1).}

The dampening effect of fluctuation-dependent coexistence on increasing
environmental variability depends on the specific traits (parameter
values) of the species in the regional pool.
\new{Under the storage effect, moderately asymmetric competition} makes
it more difficult for new species to enter the local community, but once
they do enter, ecosystem \emph{CV} is similar between communities with
low and moderate competitive asymmetries (Fig. 3; compare top and bottom
panels). Moderately asymmetric competition does decrease the rate at
which ecosystem \emph{CV} increases with environmental variance (Fig.
SI-8) because the abundance of inferior competitors is reduced and they
do not influence ecosystem \emph{CV} as much as when competitive
asymmetry is low. The correlation of species' environmental responses
(\(\rho\)) also mediates the relationship between environmental
variance, species richness, and ecosystem \emph{CV}: lower correlations
make it easier for new species to enter the community and contribute to
porfolio effects (Fig. 3). When the correlation of species'
environmental responses were as negative as possible (\(\rho = -1/3\)),
ecosystem \emph{CV} of the four-species community was immune to
increases in the environmental cue variance (Fig. 3A). However, more
extreme increases in the variance of the environmental cue, which
increase the number of extremely low or high germination events (i.e.,
\(\gamma \approx 0\) or 1; Fig. SI-9), eventually caused ecosystem
\emph{CV} to increase in the four species community (Fig. SI-10).

>>>>

MacArthur (1955), Elton (1958), and even Darwin (Turnbull et al. 2013)
recognized the potential for compensatory dynamics among species to
stabilize ecosystem functioning in fluctuating environments. This idea
underlies the ``insurance hypothesis'' (Yachi \& Loreau 1999), which
states that ecosystem variability, defined as the coefficient of
variation of ecosystem biomass over time, should decrease with diversity
because species respond dissimilarly to environmental variation,
broadening the range of conditions under which the community maintains
function (Loreau 2010). A variety of theoretical models all predict a
negative relationship between species richness and ecosystem variability
(Lehman \& Tilman 2000; Ives \& Hughes 2002; Loreau \& de Mazancourt
2013), and experimental tests tend to support such a prediction (Tilman
et al. 2006; Hector et al. 2010).

However, the ability of biodiversity--ecosystem functioning (BEF)
experiments to accurately represent real-world dynamics is debated
(Eisenhauer et al. 2016; Wardle 2016). Much of the debate centers on the
fact that BEF experimental protocols do not allow species additions from
the regional pool to offset species losses in local communities.
Theoretical work on diversity--ecosystem variability relationships
typically suffers from the same limitation: it recognizes the role of
environmental variability in driving population fluctuations which
destabilize ecosystems, but ignores the potential for environmental
variability to promote species richness and thereby help stabilize
ecosystems (Loreau 2010, but see Chesson et al. 2001).

Fluctuating environmental conditions are an important ingredient for
stable species coexistence, both in theoretical models (Chesson 2000a;
Chesson et al. 2004) and in natural communities (Cáceres 1997;
Descamps-Julien \& Gonzalez 2005; Adler et al. 2006; Angert et al.
2009). Such ``fluctuation-dependent'' coexistence emerges most easily
when species have unique environmental responses and environmental
conditions vary so that each species experiences both favorable and
unfavorable conditions, preventing competitive exclusion (Chesson
2000a). Chesson (2000) described the two temporal fluctuation-dependent
mechanisms, the storage effect and relative nonlinearity. Both
mechanisms operate when environmental variation favors different species
at different times. Under the storage effect, this happens because
species are competing for resources at different times (and escaping
competition in unfavorable periods). Under relative nonlinearity, all
species are competing for resources at the same time, but each species
alters resource availability in a way that favors its competitors. We
describe these mechanisms in more detail below (see
\emph{Materials and Methods: Consumer-resource model}).

@@

The opposing effects of environmental variability on ecosystem
variability might explain the mixed results from observational studies
on the diversity--ecosystem variability relationship. Observational
tests of the diversity--ecosystem variability relationship, which
require sampling across natural diversity gradients, have yielded
negative (Hautier et al. 2014), neutral (Valone \& Hoffman 2003; Cusson
et al. 2015), and positive (Sasaki \& Lauenroth 2011) relationships. In
a meta-analysis of diversity--ecosystem variability relationships, Jiang
\& Pu (2009) found no significant evidence for an effect of species
richness on ecosystem variability when restricting data to observational
studies in terrestrial ecosystems, perhaps because environmental
variability varies across natural diversity gradients, affecting both
richness and ecosystem variability. The idiosyncratic results of these
observational studies contrast with the consistent conclusions from
experimental and theoretical work that ignore, or control, the feedbacks
between variability and richness.

The gap between theoretical expectations and empirical results of
diversity--ecosystem variability relationships might reflect the
divergence of theory developed to explain species coexistence and theory
developed to explain diversity and ecosystem variability. In his
thorough review of the topic, Loreau (2010) cautions that ``one of the
pieces of the stability jigsaw {[}puzzle{]} that is still missing here
is the interconnection between community stability and the maintenance
of species diversity due to temporal environmental variability.'' One
reason these two disciplines have diverged is because they have focused
on different questions. Diversity--ecosystem variability studies
typically ask how ecosystem variability responds to different levels of
species richness at a given level of environmental variability (reviewed
in Kinzig et al. 2001; Loreau 2010), whereas coexistence studies ask how
species richness responds to different levels of environmental
variability (Chesson \& Warner 1981).

To reconcile these two perspectives, we extend theory on the
relationship between species richness and ecosystem variability to cases
in which species coexistence explicitly depends on environmental
fluctuations and species-specific responses to environmental conditions.
We focus on communities where coexistence is maintained by either the
temporal storage effect or relative nonlinearity using a general
consumer-resource model. We use the model to investigate two questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Does the diversity--ecosystem variability relationship remain negative
  when species coexistence is maintained by the temporal storage effect
  or relative nonlinearity?
\item
  How does increasing environmental variability impact ecosystem
  variability when coexistence depends on the storage effect or relative
  nonlinearity?
\end{enumerate}

@@

We developed a semi-discrete consumer-resource model that allows
multiple species to coexist on one resource by either the storage effect
or relative nonlinearity. In our model, the consumer can be in one of
two-states: a dormant state \(D\) and a live state \(N\). The dormant
state could represent the seed bank of an annual plant or root biomass
of a perennial plant. Transitions between \(N\) and \(D\) occur at
discrete intervals between growing seasons, with continuous-time
consumer-resource dynamics between the discrete transitions. Thus, our
model is formulated as ``pulsed differential equations'' (Pachepsky et
al. 2008; Mailleret \& Lemesle 2009; Mordecai et al. 2016). We refer to
\(\tau\) as growing seasons and each growing season is composed of \(T\)
daily time steps, indexed by \(t\) (\(t=1,2,3,\dots,T\)). The notation
\(\tau(t)\) reads as: ``day \(t\) within growing season \(\tau\).''

At the beginning of growing season \(\tau\) a season-specific fraction
(\(\gamma_{i,\tau}\)) of dormant biomass is activated as living biomass
such that \vspace{-3em}

@@

\noindent where \emph{i} indexes each species and \(\tau(0)\) denotes
the beginning of growing season \(\tau\). Live biomass at the start of
the growing season \(\left(N_{i,\tau(0)}\right)\) then serves as the
initial conditions for continuous-time consumer-resource dynamics that
are modeled as two differential equations: \vspace{-3em}

@@

At the end of the growing season (\(t=T\)), a fraction (\(\alpha_i\)) of
live biomass is stored as dormant biomass and a fraction of dormant
biomass survives (\(1-\eta_i\)) to the next growing season, giving the
following equation: \vspace{-3em}

@@

\noindent where \(\tau(T)\) denotes the end of growing season \(\tau\).
We assume remaining live biomass
\(\left(N_{i,\tau(T)}(1-\alpha_i)\right)\) dies (i.e., this is not a
closed system where all biomass must be in either \(N\) or \(D\)
states). We do not include extinction thresholds, or any other form of
demographic stochasticity, under the assumption that we are working with
abundant species with generous seed dispersal.

We assume the resource pool is not replenished within a growing season.
Resource replenishment occurs between growing seasons, and the resource
pool (\emph{R}) at the start of the growing season is
\(R_{\tau(0)} = R^+\), where \(R^+\) is a random resource pulse drawn
from a lognormal distribution with mean \(\mu(R^+)\) and standard
deviation \(\sigma(R^+)\). Taken all together, we can combine equations
1 and 4 to define the discrete transitions between live and dormant
biomass at the end of a growing season. Thus, the initial conditions for
each state (\(D, N, R\)) at the beginning of growing season \(\tau+1\)
are: \vspace{-3em}

@@

\noindent where, as above, \(\tau(T)\) denotes the end of growing season
\(\tau\) and \(\tau(0) + 1\) denotes the beginning of growing season
\(\tau+1\). The subscript (0) and superscript (200) indicates a
lognormal distribution truncated at those values to avoid extreme
resource pulses that cause computational problems. We used the function
\texttt{urlnorm} from the \texttt{Runuran} package (Leydold \& Hörmann
2015) to generate values from the truncated lognormal distribution.
Model parameters and notation are described in Table 1.

Our model does not include demographic stochasticity, which can lead to
stochastic extinction for small populations as environmental variability
increases (Boyce 1992). Previous work has shown how demographic
stochasticity and coexistence mechanisms can interact to create a weak
``humped-shape'' relationship between coexistence time and environmental
variability (Adler \& Drake 2008), because environmental variability
increases coexistence strength and the probability of stochastic
extinction simultaneously. We do not consider this potential effect here
because our focus is on large populations that would most influence
ecosystem functioning.

We limit our analysis to four-species communities because it is
exceedingly difficult to get more than four species to coexist via
relative nonlinearity without introducing another coexistence mechanism
(Yuan \& Chesson 2015). For consistency, we also constrain our focus to
four species communities under the storage effect, but our conclusions
apply to more species-rich communities (see Supporting Information
section SI.2).

@@

For the storage effect to operate, we need species-specific responses to
environmental variability, density-dependent covariance between
environmental conditions and competition (\emph{EC} covariance), and
subadditive population growth (Chesson 1994, 2000b). If these conditions
are present, all species can increase when rare and coexistence is
stable. In the storage effect, rare species increase by escaping the
effects of \emph{EC} covariance. Common species will experience greater
than average competition (\emph{C}) in environment (\emph{E}) years that
are good for them because common species cannot avoid intraspecific
competition. However, a rare species can escape intraspecific
competition and has the potential to increase rapidly in a year when the
environment is good for them but bad for the common species. \emph{EC}
covariance emerges in our model because dormant-to-live transition rates
(\(\gamma\)) are species-specific and vary through time. In a high
\(\gamma\) year for a common species, resource uptake will be above
average because combined population size will be above average. In a
year when \(\gamma\) is high for rare species and low for common
species, resource uptake will be below average because combined
population size will be below average. Subadditive population growth
buffers populations against large population decreases in unfavorable
years. It is included in our model through a dormant stage with very low
death rates, which limits large population declines in bad \emph{E}
years.

@@

To explore how fluctuation-dependent coexistence can affect the
diversity--ecosystem variability relationship, we simulated the model
with four species under two scenarios for each coexistence mechanism.
First, we allowed the variance of the environment to determine how many
species can coexist, akin to a community assembly experiment with a
species pool of four species. We simulated communities with all species
initially present across a gradient of annual resource variability for
relative nonlinearity (50 evenly-spaced values of \(\sigma_R\) in the
range {[}0, 1.2{]}) or environmental cue variability for the storage
effect (100 evenly-spaced values of \(\sigma_E^2\) in the range {[}0,
3{]}). Second, we chose parameter values that allowed coexistence of all
four species and then performed species removals at a single level of
environmental variability, akin to a biodiversity--ecosystem function
experiment. The two simulation experiments correspond to (i) sampling
ecosystem function across a natural gradient of species richness and
(ii) sampling ecosystem function across diversity treatments within a
site. We refer to the former as a ``regional'' relationship, and the
latter as a ``local'' relationship. But we do not attribute any
particular area size to ``region'', it is simply any area over which a
gradient of environmental variability exists.

To understand how increasing environmental variability will impact
ecosystem variability when coexistence is fluctuation-dependent, we
simulated the model over a range of species pool sizes and environmental
cue or resource variability. For each size of species pool (1, 2, 3, or
4 species), we simulated the model at 15 evenly-spaced levels of
environmental cue (range = {[}0.1, 2{]}) for the storage effect and 25
evenly-spaced levels of resource variability (range = {[}0, 1.2{]}) for
relative nonlinearity. We also explored the influence of asymmetries in
species' competitive abilities and correlations in species'
environmental responses within the storage effect model. We created
competitive hierarchies by making the live-to-dormant biomass fractions
(\(\alpha\)s) unequal among species. Small differences among values of
\(\alpha\) were needed to create competitive hierarchies because we
chose a relatively constrained gradient of environmental cue variance.
Larger differences among values of \(\alpha\) expand the region of
coexistence farther along a gradient environmental cue variance.

Under relative nonlinearity, species' resource response curves (Fig.
SI-5) reflect traits that determine the temporal variability of each
species' population growth. ``Stable'' species achieve maximum resource
uptake at low resource levels, but their maximum uptake rates are
modest. For these species, population responses to resource fluctuations
are buffered. ``Unstable'' species have very high maximum uptake rates,
which they only achieve when resource availability is high, leading to
large population fluctuations. The difference in the intrinsic stability
of these two kinds of species makes our simulations sensitive to initial
conditions. Therefore, we ran two sets of simulations for relative
nonlinearity: beginning with either stable or unstable species as a
reference point. For example, if species A is the most stable species
and species D is the least stable, we ran simulations where A then B
then C then D were added to the initial pool of species. We then ran
simulations with that order reversed.

All simulations were run for 5,000 growing seasons of 100 days each. We
averaged biomass over the growing season, and yearly values of
live-state biomass were used to calculate total community biomass in
each year. After discarding an initial 500 seasons to reduce transient
effects on our results, we calculated the coefficient of variation
(\emph{CV}) of summed species biomass through time, which represents
ecosystem variability, the inverse of ecosystem stability. We calculated
realized species richness as the number of species whose average biomass
was greater than 1 over the course of the simulation. In some cases,
realized species richness is less than number of species initialized for
a simulation because of competitive exclusion.

For parameters that we did not vary, we chose values that would allow
coexistence of all four species at some point along the environmental
variability gradients we simulated. Our focus is specifically on
communities where fluctuation-dependent coexistence is operating, and
making parameters increasingly asymmetric among species typically
reduced coexistence strength or made coexistence impossible (Supporting
Information section SI.3). Changes in the absolute values of parameters
also altered the strength of coexistence, but in no case did altering
parameter values change our qualitative results and conclusions
(Supporting Information section SI.3). Parameter values for specific
results are given in figure captions.

Within-season dynamics were solved given initial conditions using the
package \texttt{deSolve} (Soetaert et al. 2010) in \texttt{R} (Team
2013). \texttt{R} code for our model function is in the Supporting
Information section SI.1. All model code has been deposited on Figshare
(\emph{link}) and is available on GitHub at
\url{https://github.com/atredennick/Coexistence-Stability/releases}.

@@

When we allowed the variance of the environment to determine which of
four initial species coexisted, similar to a study across a natural
diversity gradient, we found a positive relationship between richness
and ecosystem variability, defined as the temporal \emph{CV} of total
community biomass (Fig. 2A,C). This was true for the storage effect,
where coexistence is maintained by fluctuating dormant-to-live
transition rates (\(\gamma\)), and for relative nonlinearity, where
coexistence is maintained by annual resource pulses. The relationship is
driven by the fact that increasing environmental variability increases
the strength of both coexistence mechanisms (Fig. SI-6). More variable
conditions promoted species richness, creating a positive relationship
between diversity and ecosystem variability.

When we performed species removals but held environmental variability at
a level that allows coexistence of all four species, similar to a
biodiversity--ecosystem functioning experiment, we found a negative
diversity--ecosystem variability relationship (Fig. 2B,D). Scatter
around the relationship was small for the storage effect because all
species have similar temporal variances. Regardless of species identity,
the presence of more species always stabilized ecosystem functioning
through portfolio effects. In contrast, scatter around the relationship
was larger for relative nonlinearity (Fig. 2D) because species with
different resource uptake curves had different population variances.
Depending on which species were present, two-species communities were
sometimes less variable than three-species communities. Furthermore, the
slope of the relative nonlinearity diversity--ecosystem variability
relationshp in Fig. 2D is sensitive to species' traits: the difference
among species' resource uptake determines the spread of single-species
communities along the y-axis. This means that the relationship can
become flat as species become more similar.

For the storage effect, total community \emph{CV} decreased with species
richness at a given level of environmental variability because
additional species reduced the temporal standard deviation due to
portfolio effects (Fig. SI-7). Mean biomass remained the same because
all species had the same resource uptake functions, which was necessary
to eliminate any potential effects of relative nonlinearity. Portfolio
effects under the storage effect remained strong in an eight-species
community, where total community \emph{CV} saturated after addition of
the fifth species (Fig. SI-1). For relative nonlinearity, total
community \emph{CV} decreased with species richness at a given level of
environmental variability because additional species increased mean
biomass (over-yielding) and, at higher richness (three to four species),
reduced the temporal standard deviation (Fig. SI-7). Mean biomass
increased because some species had higher growth rates (Fig. SI-5),
increasing total biomass.

To understand how much species additions might stabilize ecosystem
functioning as environmental variability increases, we simulated our
model over a range of environmental variance and species pool sizes. For
both coexistence mechanisms, realized species richness increased with
environmental variability and, in some cases, increases in richness
completely offset the effect of moderate increases in environmental
variability on ecosystem variability (Fig. 3 and 4). More species rich
communities were less variable on average and, under the storage effect,
they increased in ecosystem \emph{CV} at a slower rate than communities
with fewer species (e.g., lower slopes in log-log space; Fig. SI-8). The
buffering effect of species richness under the storage effect is also
evident in Fig. 2A because the relationship between species richness and
ecosystem \emph{CV} begins to saturate. In fact, ecosystem \emph{CV}
remains relatively constant past four species when species have
independent responses to the environment (\(\rho=0\); Fig. SI-1).

The dampening effect of fluctuation-dependent coexistence on increasing
environmental variability depends on the specific traits (parameter
values) of the species in the regional pool. Under the storage effect,
moderately asymmetric competition makes it more difficult for new
species to enter the local community, but once they do enter, ecosystem
\emph{CV} is similar between communities with low and moderate
competitive asymmetries (Fig. 3; compare top and bottom panels).
Moderately asymmetric competition does decrease the rate at which
ecosystem \emph{CV} increases with environmental variance (Fig. SI-8)
because the abundance of inferior competitors is reduced and they do not
influence ecosystem \emph{CV} as much as when competitive asymmetry is
low. The correlation of species' environmental responses (\(\rho\)) also
mediates the relationship between environmental variance, species
richness, and ecosystem \emph{CV}: lower correlations make it easier for
new species to enter the community and contribute to porfolio effects
(Fig. 3). When the correlation of species' environmental responses were
as negative as possible (\(\rho = -1/3\)), ecosystem \emph{CV} of the
four-species community was immune to increases in the environmental cue
variance (Fig. 3A). However, more extreme increases in the variance of
the environmental cue, which increase the number of extremely low or
high germination events (i.e., \(\gamma \approx 0\) or 1; Fig. SI-9),
eventually caused ecosystem \emph{CV} to increase in the four species
community (Fig. SI-10).
